---
title: "NPA long-term reef monitoring"
author: "Molly Wilson"
date: "7/20/2022"
output: 
  html_document:
    code_folding: hide
---
# {.tabset}

```{r, message = F, warning = F, echo = F}
library(tidyverse) # for all basic tools
library(here) # for linking files within folder organization
library(readxl) # for reading sheets within excel files
library(janitor) # for cleaning headings/cases etc.
library(snakecase) # for adjusting capitalization of text within data (e.g., species names)
library(lubridate) # for extracting different components of date data
library(forcats) # for reordering variable names
library(knitr) # for including tables
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Fish

What stats/graphs do we want?

* Overall abundance by group at each site/date
* Overall abundance by group over time (panel?)
* Size distributions within each group (panel)

To do list:

* Add herbivore-specific graphs?

```{r}
fish <- read_excel(here("reef monitoring", "data", "rd_fish.xlsx"), sheet = "fish") %>% # set this to location of data file or import manually
  clean_names() %>%
  mutate(date = ymd(date),
         year = year(date))

fish_groups <- c("Doctorfish", "Parrotfish", "Par (Rain, Mid, Blue)", "Damselfish", "Trigger/Box/Filefish", "Snappers", "Grunt/Porgy/Margate", "Coney", "Hind", "Nassau Grouper", "Barracuda", "Mackeral", "Jacks", "Lionfish") # names of fish groups that have size data

fish_col <- geom_col(fill = "cadetblue", color = "black", alpha = 0.5) # setting standard color for the fish 
fish_belt_area <- 30*2 # 30m x 2m belt transects
```

### Density of all fish by site and year
* This includes only fish (not sharks, turtles, etc.)
```{r}
# calculating density of all fish by site and date
fish_density_tot <- fish %>%
  filter(group %in% fish_groups) %>%
  group_by(date, year, site, transect) %>%
  summarize(count_tot = sum(count),
            density = count_tot / fish_belt_area * 100) %>%
  group_by(date, year, site) %>%
  summarize(count_mean = mean(count_tot),
            density_mean = mean(density),
            density_se = sd(density)/sqrt(n()))

ggplot(fish_density_tot, aes(x = as.character(year), y = density_mean)) +
  fish_col +
  geom_errorbar(aes(ymin = density_mean - density_se, ymax = density_mean + density_se), width = .2,
                 position = position_dodge(.9)) +
  facet_grid(. ~ site) +
  labs(x = "Year", y = expression(Density~(indv.~100~m^-2))) +
  theme_bw()
```

### Density by size, year, and site
```{r}
# calculating density of fish groups of each size bin by site and date 
fish_density_size <- fish %>%
  filter(group %in% fish_groups) %>%
  expand(nesting(site, date, year, transect), size_bin) %>% # expand df to include all possible size bins
  left_join(fish %>%
              select(site, date, transect, size_bin, count)) %>%
  mutate(count = if_else(is.na(count), 0, count),
         density = count/fish_belt_area) %>% 
  group_by(site, date, year, size_bin) %>%
  summarize(count_mean = mean(count),
            density_mean = mean(density),
            density_se = sd(density)/sqrt(n())) %>%
  mutate(size_bin = factor(size_bin, levels = c("0-5", "6-10", "11-20", "21-40", "41-60", ">60")))

ggplot(fish_density_size, aes(x = size_bin, y = density_mean)) +
  fish_col +
  geom_errorbar(aes(ymin = density_mean - density_se, ymax = density_mean + density_se), width = .2,
                 position = position_dodge(.9)) +
  facet_grid(year ~ site) +
  labs(x = "Length (cm)", y = expression(Density~(indv.~m^-2))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


### Density by group, site and year
```{r, fig.width = 12}
# calculating density of each group by site and date (regardless of size bin)
fish_density_grp <- fish %>%
  expand(nesting(site, date, year, transect), group) %>%
  left_join(fish %>%
              select(site, date, transect, group, count) %>%
              group_by(site, date, transect, group) %>%
              summarize(count = sum(count))) %>%
  mutate(count = if_else(is.na(count), 0, count), # replaces all entries not observed with 0
         density = count/fish_belt_area) %>%
  group_by(site, date, year, group) %>%
  summarize(count_mean = mean(count),
            density_mean = mean(density),
            density_se = sd(density)/sqrt(n()))

ggplot(fish_density_grp, aes(x = factor(year), y = density_mean)) +
  fish_col +
  geom_errorbar(aes(ymin = density_mean - density_se, ymax = density_mean + density_se), width = .2,
                 position = position_dodge(.9)) +
  facet_grid(site ~ group) +
  labs(x = "Year", y = expression(Density~(indv.~m^-2))) +
  theme_bw()
```

### Density by group, size, and year (one figure per site)

#### "RDAB07":
```{r, fig.width = 12}
# calculating density of fish groups of each size bin by site and date 
fish_density_grp_size <- fish %>%
  expand(nesting(site, date, year, transect), group, size_bin) %>% # expand df to include all possible groups/size bins
  left_join(fish %>%
              select(site, date, transect, group, size_bin, count)) %>%
  mutate(count = if_else(is.na(count), 0, count),
         density = count/fish_belt_area) %>% 
  group_by(site, date, year, group, size_bin) %>%
  summarize(count_mean = mean(count),
            density_mean = mean(density),
            density_se = sd(density)/sqrt(n())) %>%
  mutate(size_bin = fct_relevel(size_bin, "0-5", "6-10", "11-20", "21-40"))

ggplot(fish_density_grp_size %>% filter(site == "RDAB07"), aes(x = size_bin, y = density_mean)) +
  fish_col +
  geom_errorbar(aes(ymin = density_mean - density_se, ymax = density_mean + density_se), width = .2,
                 position = position_dodge(.9)) +
  facet_grid(year ~ group) +
  labs(x = "Length (cm)", y = expression(Density~(indv.~m^-2))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
